neonstore: provenance-aware persistent storage for NEON data

### 2021-03-23 paper/paper.Rmd
:pen:

---
title: "neonstore: a local, provenance-aware workflow for NEON data"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "Quinn Thomas"
    affiliation: vt
  - name: "Christine Laney"
    affiliation: neon
  - name: "Claire Lunch"
    affiliation: neon
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: preprint
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.  The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.  `neonstore` design emphasizes simplicity and data provenance. Here, we describe the design, use, and rationale for the package.

@@

NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.  Meanwhile, products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need. As some NEON data products can be quite large (NEON's data catalog to date is in excess of 1 petabyte, primarily due to the size of many AOP products), being able to only download or only read into R the files of interest is already important to save bandwidth or computer memory, and this will be increasingly true as NEON data accumulates in the years ahead. This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product. Users can easily verify whether or not the update has changed data in their analyses by comparing file hashes, rather than having to read in and examine data meticulously row by row.

However, the highly atomized structure of NEON's data can also become a barrier to entry without convenient tools for working with many files.  Most analyses will inevitably require multiple files to be combined or stacked, e.g. to create a unified record for a product across many months or years of sampling or across many sites.  When the process of combining files is complex or slow to execute, many researchers will resort to saving the resulting 'stacked' products as a basis for future work.  Such workflows may be more cumbersome to update as additional data continues to be published, or to compare against any subsequent corrections to previously-published data.  `neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.

>>>>

---
title: "neonstore: provenance-aware persistent storage for NEON data"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "Quinn Thomas"
    affiliation: vt
  - name: "Christine Laney"
    affiliation: neon
  - name: "Claire Lunch"
    affiliation: neon
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: preprint
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.  The `neonstore` R package seeks to provide provenance aware, high performance access and persistent storage of NEON data files. Here, we describe the design, use, and rationale for the package.

@@

Provenance and performance are two central aspects to the design of `neonstore`.  A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.  A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available RAM on most machines (e.g. 10 GB - 10 TB range, depending on products and sites considered).

NEON data products are distributed using a highly atomized file design: with each data product from either both observation systems (OS) and instrument systems (IS) divided into separate files by site and month.  Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need. As some NEON data products can be quite large (NEON's complete data catalog to date is already in excess of 1 petabyte, primarily due to the size of many AOP products), being able to download or read only the files of interest is already important to save bandwidth or computer memory, or disk space. This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product. However, the highly atomized structure of NEON's data can also become a barrier to entry.  Most analyses will inevitably require multiple files to be combined or stacked, e.g. to create a unified record for a product across many months or years of sampling or across many sites. In addition to releasing new data files each month, NEON will occasionally update previously released data files.  More confusingly, NEON frequently updates the timestamp associated with a file even when the file has not changed. Typically these events are associated with the update of a separate but related file.

Any analysis is only as good as as the data supporting it.  Because NEON data is subject to both revisions (where previously released data are corrected) and new releases (where more recently collected data is made public), researchers will seek to update previous analyses with corrected data and new observations.  `neonstore` enables scripted workflows through which such updates can occur automatically by re-running the same script.  However, when new or revised data significantly alter results, it is also important to be able to pinpoint precisely what has changed and why.  These elements creates five distinct tasks for managing NEON data provenance: (1) The ability to determine which of each of the individual raw data files were used in an analysis (2) The ability to append new data as it is collected and released (3) the ability to transparently replace existing data when corrections are published, (4) the ability to avoid re-downloading or re-stacking data that has been previously processed, (5) the ability to freeze the data used in a given analysis, so that results can reliably be repeated without accidentally including revised or newer data products.

To provide these five abilities for data provenance, `neonstore` maintains a local file store containing all raw data files downloaded from NEON. Additionally, each file is entered into a local registry, along with the file's MD5 checksum (CRC32 for AOP products) and currently associated NEON release (if any).  This local registry is kept in a Lightning Memory Mapped Database, LMDB, an extremely fast, open source key-value store indexed on both file name and checksum.  MD5 checksums effectively indicate unique content. It is theoretically possible for a malicious actor to create a file with different contents but the same checksum, but vanishingly unlikely to occur by chance. On requesting new files to download, `neonstore` compares file names and hashes reported by the API to those in the local store, allowing it to omit any data that has already been downloaded, even if the timestamp in a file name has changed but the contents remain the same.

NEON's filenaming convention provided essential metadata about each file, such as the data product (product level, product number, revision number), site, year and month of the release, file description, a timestamp.  Sensor data (IS) also include additional information about the horizontal and vertical position and meaasurement interval (frequency) of the sensor product.  AOP products use a somewhat different convention with flight information.  `neonstore` recognizes all published NEON file naming conventions and automatically parses NEON filenames into their component parts.  Data are organized into subdirectories based on product, site, and year-month respectively.  Data or metadata files shared across all months or all sites of a product are stored in the corresponding parent directory. The function `neon_index()` returns a `data.frame` listing every recognized individual file in the local store, along with the corresponding metadata fields parsed from the filename. Optional arguments can be given to restrict this table to files matching specific products, sites, date ranges, etc.

Box 1: Provenance workflow
- Existing NEON data used in an analysis to support a result.
- A data file is corrected for an error. Revised data file is downloaded, imported into `neonstore` database, replacing deprecated data. Deprecated data file is retained in local file store for future reference.  Analysis confirms same result
- Next month's data are released, downloaded, imported by `neonstore`, and then used to revise model and provide updated result

`neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.
